---
title: "Lab 11 - Grading the professor, Pt. 2"
subtitle: "Modeling with multiple predictors"
author: "Your Name"
output: 
  tufte::tufte_handout:
    latex_engine: xelatex
    highlight: pygments
    keep_tex: true
  tufte::tufte_html:
    css: ../lab.css
    tufte_variant: "envisioned"
    highlight: pygments
link-citations: yes
---

```{r include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Introduction

In this lab we revisit the professor evaluations data we modeled in Lab 10. In the last lab we modeled evaluation scores using a single predictor at a time. This time we will use multiple predictors to model evaluation scores.

**This lab builds directly on Lab 10.** Make sure you understand single predictor models before working with multiple predictors.

For full context, review Lab 10's introduction before continuing.

**Estimated time:** 60-90 minutes

## Learning Goals

By the end of this lab, you will be able to:

-   Fit linear regression models with multiple predictors
-   Interpret regression coefficients in multiple regression
-   Compare simple and multiple regression models
-   Understand adjusted R² and when to use it
-   Interpret interaction effects between predictors
-   Make predictions using multiple regression models

## Prerequisites

Before you begin, make sure you have:

-   ✅ **Completed Lab 10** (this is essential!)
-   ✅ Watched lectures on multiple regression
-   ✅ Understand single predictor regression
-   ✅ Understand the concepts of R² and adjusted R²
-   ✅ Forked the lab-instructions repository

------------------------------------------------------------------------

# Getting Started

Navigate to your forked `lab-instructions` repository in JupyterHub, open the `Lab11` folder, and open the R Markdown document `lab-11.Rmd`.

## Verify Your Setup

Before proceeding:

Step 1. Open `lab-11.Rmd` in RStudio

Step 2. Check that the Git pane shows YOUR username (not the course organization)

Step 3. Click **Knit** to make sure the document compiles

## Warm Up

**Step 1:** Update the YAML, changing the author name to your name.

**Step 2:** Knit the document.

**Step 3:** Commit your changes with message "Updated author name".

**Step 4:** Push to GitHub and verify the changes are visible in your repo.

🧶 ✅ ⬆️ **Knit, commit with message "Completed warm up", and push your changes.**

------------------------------------------------------------------------

## Packages

We'll use the **tidyverse** package for much of the data wrangling and visualization, the **tidymodels** package for modeling and inference, and the data lives in the **openintro** package. These packages are already installed for you. You can load them by running the following in your Console:

```{r load-packages, eval = TRUE, message = FALSE}
library(tidyverse) 
library(tidymodels)
library(openintro)
```

## Data

The data can be found in the **openintro** package, and it's called `evals`. Since the dataset is distributed with the package, we don't need to load it separately; it becomes available to us when we load the package. You can find out more about the dataset by inspecting its documentation, which you can access by running `?evals` in the Console or using the Help menu in RStudio to search for `evals`. You can also find this information [here](https://www.openintro.org/data/index.php?data=evals).

------------------------------------------------------------------------

# Exercises

## Exercise 1

Load the data by including the appropriate code in your R Markdown file.

**Note:** Since `evals` comes from the openintro package, it's automatically available when you load the package. However, it's good practice to verify the data loaded correctly.

```{r load-data, eval = TRUE}
# The data is already available, but let's verify it loaded
glimpse(evals)
```

**How many observations are in the dataset?** Use inline code to report.

**Your answer:** The dataset has \_\_\_\_\_ observations.

**✓ Checkpoint:** You should see 463 observations and 23 variables.

------------------------------------------------------------------------

## Part 1: Simple Linear Regression (Review)

## Exercise 2

Fit a linear model (one you have fit in Lab 10): `score_bty_fit`, predicting average professor evaluation `score` based on average beauty rating (`bty_avg`) only. Write the linear model, and note the R² and the adjusted R².

```{r fit-score-bty, eval = FALSE}
# Remember to change eval=FALSE to eval=TRUE!
score_bty_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(score ~ ___, data = evals)

# View coefficients
tidy(score_bty_fit)

# View model statistics including R²
glance(score_bty_fit)
```

**Write the linear model equation:**

$$\widehat{score} = \_\_\_ + \_\_\_ \times bty\_avg$$

**R² value:** \_\_\_\_\_

**Adjusted R² value:** \_\_\_\_\_

🧶 ✅ ⬆️ **Knit, commit with message "Completed Exercises 1-2", and push your changes.**

------------------------------------------------------------------------

## Part 2: Multiple Linear Regression

## Exercise 3

Fit a linear model (one you have fit in Lab 10): `score_bty_gen_fit`, predicting average professor evaluation `score` based on average beauty rating (`bty_avg`) and `gender`. Write the linear model, and note the R² and the adjusted R².

```{r fit-score-bty-gender, eval = FALSE}
# Remember to change eval=FALSE to eval=TRUE!
score_bty_gen_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(score ~ ___ + ___, data = evals)

# View coefficients
tidy(score_bty_gen_fit)

# View model statistics
glance(score_bty_gen_fit)
```

**Write the linear model equation:**

$$\widehat{score} = \_\_\_ + \_\_\_ \times bty\_avg + \_\_\_ \times gendermale$$

**R² value:** \_\_\_\_\_

**Adjusted R² value:** \_\_\_\_\_

------------------------------------------------------------------------

## Exercise 4

Interpret the slopes and intercept of `score_bty_gen_fit` in context of the data.

**Interpret the intercept:**

The intercept represents

**Interpret the slope for bty_avg:**

Holding gender constant, for every one unit increase in beauty rating, \_\_\_\_\_\_\_

**Interpret the slope for gendermale:**

Holding beauty rating constant, male professors are expected to score \_\_\_\_\_\_\_\_

**Important:** In multiple regression, we interpret each coefficient "holding all other variables constant."

------------------------------------------------------------------------

## Exercise 5

What percent of the variability in `score` is explained by the model `score_bty_gen_fit`?

**Use the R² value from Exercise 3.**

Approximately \_\_\_\_\_% of the variability in professor evaluation scores is explained by beauty rating and gender.

------------------------------------------------------------------------

## Exercise 6

What is the equation of the line corresponding to *just* male professors?

**Hint:** For male professors, gendermale = 1.

**Starting with:** $\widehat{score} = b_0 + b_1 \times bty\_avg + b_2 \times gendermale$

**For males (gendermale = 1):**

$$\widehat{score} = b_0 + b_1 \times bty\_avg + b_2 \times 1$$

**Simplified:**

$$\widehat{score} = (b_0 + b_2) + b_1 \times bty\_avg$$

**With your actual values:**

$$\widehat{score} = \_\_\_ + \_\_\_ \times bty\_avg$$

🧶 ✅ ⬆️ **Knit, commit with message "Completed Exercises 3-6", and push your changes.**

------------------------------------------------------------------------

## Exercise 7

For two professors who received the same beauty rating, which gender tends to have the higher course evaluation score?

**Look at the coefficient for gendermale.** Is it positive or negative?

**Your answer:**

If the coefficient is positive: Male professors tend to have higher scores. If the coefficient is negative: Female professors tend to have higher scores.

**The coefficient is:** \_\_\_\_\_ (positive/negative)

**Therefore:**

**By how much?** \_\_\_\_\_

------------------------------------------------------------------------

## Exercise 8

How does the relationship between beauty and evaluation score vary between male and female professors?

**Look at the coefficient for bty_avg in the multiple regression model.**

**Your answer:**

The relationship between beauty and evaluation score \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_ (does / does not) vary between male and female professors because \_\_\_\_\_\_\_\_\_\_

**Hint:** In this model, both genders have the same slope for bty_avg. The lines are parallel!

------------------------------------------------------------------------

## Exercise 9

How do the adjusted R² values of `score_bty_gen_fit` and `score_bty_fit` compare? What does this tell us about how useful `gender` is in explaining the variability in evaluation scores when we already have information on the beauty score of the professor?

**Compare the adjusted R² values:**

**score_bty_fit adjusted R²:** \_\_\_\_\_ (from Exercise 2)

**score_bty_gen_fit adjusted R²:** \_\_\_\_\_ (from Exercise 3)

**Difference:** \_\_\_\_\_

**Interpretation:**

The adjusted R² increased/decreased by \_\_\_\_\_ when we added gender to the model. This tells us that gender \_\_\_\_\_\_

**Note:** Adjusted R² accounts for the number of predictors, so it's better for comparing models with different numbers of predictors.

------------------------------------------------------------------------

## Exercise 10

Compare the slopes of `bty_avg` under the two models (`score_bty_fit` and `score_bty_gen_fit`). Has the addition of `gender` to the model changed the parameter estimate (slope) for `bty_avg`?

**Slope of bty_avg in score_bty_fit:** \_\_\_\_\_ (from Exercise 2)

**Slope of bty_avg in score_bty_gen_fit:** \_\_\_\_\_ (from Exercise 3)

**Has it changed?**

The slope has changed/not changed. The difference is \_\_\_\_\_.

**Why might this happen?**

**Hint:** If beauty and gender are related (e.g., if male and female professors have different average beauty ratings), adding gender can change the bty_avg coefficient.

🧶 ✅ ⬆️ **Knit, commit with message "Completed Exercises 7-10", and push your changes.**

------------------------------------------------------------------------

## Exercise 11

Create a new model called `score_bty_rank_fit` with `gender` removed and `rank` added in. Write the equation of the linear model and interpret the slopes and intercept in context of the data.

```{r fit-score-bty-rank, eval = FALSE}
# Remember to change eval=FALSE to eval=TRUE!
score_bty_rank_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(score ~ ___ + ___, data = evals)

# View coefficients
tidy(score_bty_rank_fit)

# View model statistics
glance(score_bty_rank_fit)
```

**Note:** `rank` has three levels: teaching (reference), tenure track, and tenured.

**Write the linear model equation:**

$$\widehat{score} = \_\_\_ + \_\_\_ \times bty\_avg + \_\_\_ \times ranktenuretrack + \_\_\_ \times ranktenured$$

**Interpret the intercept:**

For teaching faculty with a beauty rating of 0, the expected evaluation score is \_\_\_\_\_.

**Interpret the slope for bty_avg:**

Holding rank constant, for every one unit increase in beauty rating, \_\_\_\_\_\_\_\_

**Interpret the slope for ranktenuretrack:**

Holding beauty rating constant, tenure track faculty are expected to score \_\_\_\_

**Interpret the slope for ranktenured:**

Holding beauty rating constant, tenured faculty are expected to score \_\_\_\_\_\_\_\_

**Adjusted R² for this model:** \_\_\_\_\_

**How does this compare to score_bty_gen_fit?**

🧶 ✅ ⬆️ **Knit, commit with message "Completed Exercise 11", and push your changes.**

------------------------------------------------------------------------

## Common Errors and Troubleshooting

**Multiple regression errors:**

-   **Coefficients look wrong** → Check that you included the `+` between predictors
-   **Can't interpret "holding other variables constant"** → This means we compare observations that have the same value for other predictors
-   **Confused about which R² to use** → Use adjusted R² when comparing models with different numbers of predictors

**Model comparison errors:**

-   **Can't see difference in R²** → Look at more decimal places in `glance()` output
-   **Slopes changed a lot** → This can happen if predictors are correlated
-   **Not sure which model is better** → Higher adjusted R² generally means better, but also consider context and interpretability

**Interpretation errors:**

-   **Forgot "holding other variables constant"** → This is crucial in multiple regression!
-   **Confused about categorical variables** → Remember the reference level; coefficients are relative to that
-   **Intercept doesn't make sense** → That's okay if it represents values outside the data range (like beauty = 0)

**General issues:**

-   **Code chunk not running** → Make sure you changed `eval=FALSE` to `eval=TRUE`
-   **Can't remember coefficients** → Run `tidy(model_name)` again
-   **Need more decimal places** → Use `tidy(model_name) %>% print(n = Inf, width = Inf)`

------------------------------------------------------------------------

## Reflection

**Key differences between simple and multiple regression:**

-   **Simple regression:** One predictor, one slope to interpret
-   **Multiple regression:** Multiple predictors, interpret each "holding others constant"
-   **R² vs. Adjusted R²:** Adjusted R² penalizes adding unnecessary predictors

**Important concepts:**

-   Coefficients in multiple regression show the effect of one predictor while controlling for others
-   Adding predictors can change existing coefficients if predictors are correlated
-   Adjusted R² is better for model comparison than regular R²
-   Higher adjusted R² doesn't always mean a "better" model - consider interpretability and context

------------------------------------------------------------------------

**To submit to Canvas:**

Step 1. In RStudio, click the **Knit** dropdown menu (next to the Knit button)

Step 2. Select **Knit to tufte_handout** to generate a PDF

Step 3. Download the PDF file from the Files pane

Step 4. Upload the PDF to Canvas

**✓ Final Checkpoint:** Visit your GitHub repo one more time to confirm all your work is there. We will grade what we see in your repo on GitHub!
